White Dwarf Mergers on Adaptive Meshes\\ II. Collisions

### 2017-07-09 papers/wdmerger_II/paper.tex
Update names, URLs

% Software names
\newcommand{\boxlib}{\texttt{BoxLib}}
\newcommand{\castro}{\texttt{CASTRO}}
\newcommand{\microphysics}{\texttt{Microphysics}}
\newcommand{\wdmerger}{\texttt{wdmerger}}
\newcommand{\python}{\texttt{Python}}
\newcommand{\matplotlib}{\texttt{matplotlib}}
\newcommand{\yt}{\texttt{yt}}
\newcommand{\vode}{\texttt{VODE}}
\newcommand{\isoseven}{\texttt{iso7}}
\newcommand{\aproxthirteen}{\texttt{aprox13}}
\newcommand{\aproxnineteen}{\texttt{aprox19}}
\newcommand{\aproxtwentyone}{\texttt{aprox21}}

@@

In this paper we seek to add another view to the problem using the reactive hydrodynamics
code \castro\ \citep{castro} and its underlying adaptive mesh refinement library \boxlib.
This paper is a follow-up to \citet{wdmergerI} (hereafter, Paper I), which established
the hydrodynamic methdology we are using to simulate mergers of white dwarfs.
Paper I left out a discussion of the methodology we are using for nuclear reactions,
and we fill out this information below in \autoref{sec:numericalmethods}. While it is
our aim is to add information to the discussion about collisions of white dwarfs,
we also view this work as a methodology study that asks to what extent we
can trust the reliability of explosive thermonuclear reactions in Eulerian
hydrodynamics simulations. As a corollary, to what extent does static and/or
adaptive mesh refinement resolve the types of problems that occur in low resolution
simulations? We discuss this in Section \autoref{sec:resolution}, though first
we examine the response of the code to a number of user-specified parameters in
\autoref{sec:parameters}. We then close with our thoughts on what our work implies
for the broader astrophysical reacting hydrodynamics literature in \autoref{sec:conclusion}.

@@

All four of these networks have been ported into a form that is consistent
with the \boxlib\ codes, in the freely available \microphysics\ code
repository\footnote{\microphysics\ can be obtained at \url{https://github.com/BoxLib-Codes/Microphysics}.},
a collection of microphysical routines that are designed to be used in our
hydrodynamics codes. These can be easily swapped at compile time by using the 
appropriate makefile variable.

@@

In our \microphysics\ repository we provide several software options for
solving a set of coupled stiff ODEs. For this work we use an implementation of
the well known variable-order Richardson extrapolation method presented by \citet{stoer:1980},
that is similar to the integrator which ships with
the original versions of the networks mentioned above. Previous work using the \boxlib\
codes typically used the classic \vode\ integrator \citep{vode}. We do maintain a version
of the software that is compatible with our software interfaces in the \microphysics\
repository, but we have largely shifted to integrators which are written in modern
Fortran as a consequence of our efforts to run our codes on GPUs. The Stoer and Bulirsch
integrator we provide satisfies this criterion; we also provide a rewrite of the VODE
BDF algorithm that uses modern Fortran features. The Stoer and Bulirsch algorithm
relies on a uniform relative error tolerance for all of the ODEs in the system, which
we set at $10^{-6}$ for all simulations described here.

@@

We agree with \citet{kushnir:2013} that solving this numerical
instability is crucial to avoiding unphysical detonations.
A simulation that does not solve this problem will not obtain
the correct amount and will not converge properly with resolution.
This may justify the addition of many AMR levels to the domain
if a correct evaluation of the burning phase is desired.
Even if one cannot afford the full resolution required by this
AMR criterion, and chooses to limit the number of levels to some
predetermined maximum based on a constraint of computing time,
the added resolution will still go to the most-needed places.
However, a significant drawback to this approach is that it
does not turn on until after an unstable burning region has
been generated, so this AMR criterion cannot help in the case
where a spurious detonation begins in a single timestep on the coarse
grid, though it will kick in immediately after that timestep
to resolve the regions where the detonation has occurred.
A possible remedy to be explored in future work is to add
refinement pre-emptively in regions where the ratio $\tau_s / \tau_e$
approaches $f_{s}$ but does not yet exceed it. Another
possibility for this case would be to track how many levels
would have been needed to prevent the numerical detonation,
based on the stability criterion, and reject the timestep and
retry it from the starting state using that many levels of refinement.
As mentioned in \autoref{sec:burnerhydrocoupling}, the ability to reject
a timestep and retry it is functionality that
we have recently added to \castro, so we may try this in future work.

@@

\software{BoxLib\footnote{\url{https://github.com/BoxLib-Codes/BoxLib}},
          CASTRO \citep{castro} (https://github.com/BoxLib-Codes/Castro),
          wdmerger \citep{wdmergerI} (\url{https://github.com/BoxLib-Codes/wdmerger}),
          GCC (\url{https://gcc.gnu.org/}),
          python (\url{https://www.python.org/}),
          matplotlib \citep{matplotlib}, \url{http://matplotlib.org/}),
          yt \citep{yt}, \url{http://yt-project.org/})}

>>>>

% Software names
\newcommand{\amrex}{\texttt{AMReX}}
\newcommand{\boxlib}{\texttt{BoxLib}}
\newcommand{\castro}{\texttt{CASTRO}}
\newcommand{\microphysics}{\texttt{Microphysics}}
\newcommand{\wdmerger}{\texttt{wdmerger}}
\newcommand{\python}{\texttt{Python}}
\newcommand{\matplotlib}{\texttt{matplotlib}}
\newcommand{\yt}{\texttt{yt}}
\newcommand{\vode}{\texttt{VODE}}
\newcommand{\isoseven}{\texttt{iso7}}
\newcommand{\aproxthirteen}{\texttt{aprox13}}
\newcommand{\aproxnineteen}{\texttt{aprox19}}
\newcommand{\aproxtwentyone}{\texttt{aprox21}}

@@

In this paper we seek to add another view to the problem using the reactive hydrodynamics
code \castro\ \citep{castro} and its underlying adaptive mesh refinement library \amrex\footnote{\amrex is the successor to \boxlib.}.
This paper is a follow-up to \citet{wdmergerI} (hereafter, Paper I), which established
the hydrodynamic methdology we are using to simulate mergers of white dwarfs.
Paper I left out a discussion of the methodology we are using for nuclear reactions,
and we fill out this information below in \autoref{sec:numericalmethods}. While it is
our aim is to add information to the discussion about collisions of white dwarfs,
we also view this work as a methodology study that asks to what extent we
can trust the reliability of explosive thermonuclear reactions in Eulerian
hydrodynamics simulations. As a corollary, to what extent does static and/or
adaptive mesh refinement resolve the types of problems that occur in low resolution
simulations? We discuss this in Section \autoref{sec:resolution}, though first
we examine the response of the code to a number of user-specified parameters in
\autoref{sec:parameters}. We then close with our thoughts on what our work implies
for the broader astrophysical reacting hydrodynamics literature in \autoref{sec:conclusion}.

@@

All four of these networks have been ported into a form that is consistent
with the \amrex\ codes, in the freely available \microphysics\ code
repository\footnote{\microphysics\ can be obtained at \url{https://github.com/starkiller-astro/Microphysics}.},
a collection of microphysical routines that are designed to be used in our
hydrodynamics codes. These can be easily swapped at compile time by using the 
appropriate makefile variable.

@@

In our \microphysics\ repository we provide several software options for
solving a set of coupled stiff ODEs. For this work we use an implementation of
the well known variable-order Richardson extrapolation method presented by \citet{stoer:1980},
that is similar to the integrator which ships with
the original versions of the networks mentioned above. Previous work using the \amrex\
codes typically used the classic \vode\ integrator \citep{vode}. We do maintain a version
of the software that is compatible with our software interfaces in the \microphysics\
repository, but we have largely shifted to integrators which are written in modern
Fortran as a consequence of our efforts to run our codes on GPUs. The Stoer and Bulirsch
integrator we provide satisfies this criterion; we also provide a rewrite of the VODE
BDF algorithm that uses modern Fortran features. The Stoer and Bulirsch algorithm
relies on a uniform relative error tolerance for all of the ODEs in the system, which
we set at $10^{-6}$ for all simulations described here.

@@

We agree with \citet{kushnir:2013} that solving this numerical
instability is crucial to avoiding unphysical detonations.
A simulation that does not solve this problem will not obtain
the correct amount and will not converge properly with resolution.
This may justify the addition of many AMR levels to the domain
if a correct evaluation of the burning phase is desired.
Even if one cannot afford the full resolution required by this
AMR criterion, and chooses to limit the number of levels to some
predetermined maximum based on a constraint of computing time,
the added resolution will still go to the most-needed places.
However, a significant drawback to this approach is that it
does not turn on until after an unstable burning region has
been generated, so this AMR criterion cannot help in the case
where a spurious detonation begins in a single timestep on the coarse
grid, though it will kick in immediately after that timestep
to resolve the regions where the detonation has occurred.
A possible remedy to be explored in future work is to add
refinement pre-emptively in regions where the ratio $\tau_s / \tau_e$
approaches $f_{s}$ but does not yet exceed it. Another
possibility for this case would be to track how many levels
would have been needed to prevent the numerical detonation,
based on the stability criterion, and reject the timestep and
retry it from the starting state using that many levels of refinement.
As mentioned in \autoref{sec:hydrocoupling}, the ability to reject
a timestep and retry it is functionality that
we have recently added to \castro, so we may try this in future work.

@@

\software{\amrex\ (\url{https://github.com/AMReX-Codes/Amrex}),
          \castro\ \citep{castro} (\url{https://github.com/AMReX-Astro/Castro}),
          \wdmerger\ \citep{wdmergerI} (\url{https://github.com/AMReX-Astro/wdmerger}),
          GCC (\url{https://gcc.gnu.org/}),
          python (\url{https://www.python.org/}),
          matplotlib \citep{matplotlib} (\url{http://matplotlib.org/}),
          yt \citep{yt} (\url{http://yt-project.org/})}
